---
title: "Predicting Drivers of Common Buckthorn"
author: Sean Colby
subtitle: Generation of a Buckthorn SDM
---

# Introduction

Common Buckthorn

Background: 
Common Buckthorn is a deciduous tree/shrub species which was introduced in the late 1800's as an ornamental from Eurasia (Knight et al. 2007). While it may have served a role as an ornamental, it has done a large amount of ecological harm to the Northeastern United States and South East Canada. For example, Common Buckthorn has been known to more than double nitrogen and organic matter content in soils compared to native vegetation resulting in decreased habitat diversity (Klionsky et al. 2011). Furthermore, since Buckthorn is highly competitive having a higher growth rate and fecundity than most native shrubs, it is then able to become quickly become a monoculture and out shade native species. To make matters worse, Buckthorn is allelopathic and reduces native species richness and native biomass production (Warren et al. 2017). 


Problem:
Due to Common Buckthorn being extremely damaging to native ecosystems it is imperative to model its spread in order to attempt to control it. While many researchers have done large multiple state prediction models of Common Buckthorn, to my knowledge very few have done a small scale single state SDM.


Questions: 
1. Will this small scale of an SDM be relatively accurate? 
2. What can we make of the plot as to what forests/areas appear to be more likely to be invaded than others?
3. Since we are focusing on habitat suitability, is buckthorn's presence more likely due to habitat suitability or dispersal methods?


# Materials and methods

Narrative: Clear narrative description of the data sources and methods. Includes data from at least two sources that were integrated / merged in R.

Code: The code associated with the project is well organized and easy to follow. Demonstrates mastery of R graphics and functions.

Data: The underlying data are publicly accessible via the web and downloaded/accessed within the Rmd script. If you want to use your own data, you must make it available on a website (e.g. Figshare) so that others are able to re-run your code.

You can do bullets like this:

* The first most important thing
* The second most important thing
* The third most important thing

You can do numbers like this:

1. The first most important thing
2. The second most important thing
3. The third most important thing

See [http://rmarkdown.rstudio.com/](http://rmarkdown.rstudio.com/) for all the amazing things you can do.


Here's my first code chunk.
```{r}
1+2
```

Load any required packages in a code chunk (you may need to install some packages):

```{r, message=F, warning=F}
library(tidyverse)
library(leaflet)
library(kableExtra)
knitr::opts_chunk$set(cache=TRUE)  # cache the results for quick compiling
```

## Download and clean all required data
```{r}
n=20
data=data.frame(x=runif(n,-180,180),
                y=runif(n,-60,60),
                size = runif(n, 5, 20),
                category = factor(
                  sample(letters[1:5], n, replace = TRUE)
                  ),
                value = rnorm(n))
```

```{r, results='asis'}
data %>% 
  slice(1:10) %>% #show only 1:n rows
  kable(digits=2,align="c")%>% #make table and round to two digits
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive")) #apply other formatting
```

Add any additional processing steps here.

# Results

[~200 words]

Tables and figures (maps and other graphics) are carefully planned to convey the results of your analysis. Intense exploration and evidence of many trials and failures. The author looked at the data in many different ways before coming to the final presentation of the data.

Show tables, plots, etc. and describe them.

```{r, fig.width=6, fig.height=3, fig.cap="Map of completely random data"}
m <- leaflet(data) %>% 
  addTiles() %>% 
  addCircleMarkers(~x, ~y, radius = ~size,color = ~as.factor(category)) %>% 
  addPopups(~x[2], ~y[2], "Random popup")
m  # a map with the default OSM tile layer
```


```{r}
data %>% 
  ggplot(aes(x=x,y=y,col=category))+
  geom_point()
```

# Conclusions

[~200 words]

Clear summary adequately describing the results and putting them in context. Discussion of further questions and ways to continue investigation.

# References

All sources are cited in a consistent manner
